---
title: "Readings"
output: blogdown::html_page
menu:
  docs:
    parent: "syllabus"
weight: 150
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Textbook

The primary reference book for this course is [R for Data Science, 2nd edition](https://r4ds.hadley.nz), by Wickham, Cetinkaya-Rundel, & Grolemund (2023).

Class sessions, and the examples worked in class and on problem sets, will not strictly follow the textbook in chronological order. There are no "required" readings. Nonetheless, if you like using reference texts to supplement your work, the textbook is an excellent reference for reading more about how and why to use the techniques we practice in class.

## Additional references

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "As we go through the class, this section will be updated with links to useful readings and resources.")
```

### Other introductory data science textbooks

- [Learning Data Science, 1st edition](https://learningds.org/intro.html), by Lau, Gonzalez, & Nolan (2023): More technical than R for Data Science, and mainly uses Python for the examples

### Package documentation & use guides

- [Tidyverse online package documentation](https://www.tidyverse.org/packages/): much more detailed and comprehensive
- [Tidyverse cheatsheets](https://rstudio.github.io/cheatsheets/): quicker one-stop-shop help, but less detailed
- [R Graphics Cookbook](https://r-graphics.org) by Winston Chang, which uses ggplot2 (which we use)


### Research guides

- [Carleton College Library research guide](https://gouldguides.carleton.edu/c.php?g=146834&p=964943) on evaluating the quality of a dataset

### Theory papers

[_Tidy data_](https://vita.had.co.nz/papers/tidy-data.pdf) by Hadley Wickham, Journal of Statistical Software
