---
title: "Overview"
output: blogdown::html_page
menu:
  docs:
    parent: "schedule"
weight: 200
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
blogdown::shortcode_html(.name = "alert",
                         icon = "‼️",
                         .content = "The schedule is subject to change, but whatever you see here is current.")
```

## Week 1

**Introduction to scientific computing**

### Day 1

- Why scientists program
- The anatomy of code and code editors
- What emotions are and how we study them

### Day 2

- Data types
- Logical and relational operators
- Tidy dataframe manipulation
- Troubleshooting as an experimentation skill

## Week 2

**Data wrangling: The core toolkit**

### Day 1

- The structure of psychology research data
- Cleaning data

### Day 2

- Reshaping data
- Joining data
- Dealing with irregular data

## Week 3

**Code style and hygiene**

### Day 1

- R scripts and R Markdown
- Scripting hygiene
- Version control with git

### Day 2

**Working with scientists' data**

- How to search for literature and datasets
- Inspecting and evaluating datasets
- Basic descriptive statistics

## Week 4

**Graphing and visual communication**

### Day 1

- The grammar of graphics
- Graphs as communication devices

### Day 2

- Graph customization
- Visual communication best practices

## Week 5

**Complex descriptive statistics and visualizations**

### Day 1

Dimension reduction and mapping

- Principal components analysis
- Multidimensional scaling
- t-SNE

### Day 2

Clustering

- Hierarchical clustering
- K-means clustering

## Week 6

**Collaborating on code and science**

### Day 1

- Collaboration with GitHub
- Peer reviewing code effectively and respectfully

### Day 2

- Preparing and documenting code and data for sharing
- Formatting and publishing analysis write-ups
