---
title: "Class notes: June 15, 2023"
output: blogdown::html_page
description: "Lecture notes on (some of) the core data manipulation functions."
excerpt: "Lecture notes on (some of) the core data manipulation functions."
date: 2023-06-15
lastmod: "`r Sys.Date()`"
draft: false
images: []
categories: ["Class notes"]
tags: ["2023"]
contributors: ["Monica Thieu"]
pinned: false
homepage: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Lecture notes

This isn't the exact _order_ that we practiced these concepts in class, but I've reordered these into an order that reads/flows more sensibly after the fact.

You can read more about these in [Chapter 4 of R for Data Science](https://r4ds.hadley.nz/data-transform.html). For more technical, in-depth information about dataframe manipulation tools, check out the package docs for [dplyr,](https://dplyr.tidyverse.org) the package these functions come from.

First, we need to read in the data used for the demonstrations below.

```{r}
big5_raw <- read_tsv(here::here("static", "datasets", "BIG5_20140518", "data.csv"))
```

### `|>`: the pipe

The pipe does one simple, but key, thing: **takes the object on the left-hand side and feeds it into the first argument of the function on the right-hand side.** This means that:

`a |> foo()` is equivalent to `foo(a)`. Fine and good.

`a |> foo() %>% bar(arg = TRUE)` is equivalent to `bar(foo(a), arg = TRUE)`. Now, nested function calls read left-to-right!

Generally, a block of code like:

```r
df_new <- df_old |> 
  foo() |> 
  bar(arg = TRUE) |> 
  baz()
```

is equivalent to:

```r
df_new <- baz(bar(foo(df_old), arg = TRUE))
```

Now, you can chain a series of commands to operate on a dataframe all at once, and easily read those commands from left-to-right and top-to-bottom, in the same order that R executes them.

In order to use a function in a |> chain, the function must:

* take as their first argument the object (usually a dataframe) to be operated upon
* return the same object, but with the operation completed

Essentially all the functions from the `tidyverse` package that we will use are pipe-safe, but bear this in mind when trying to use functions from other packages in pipe chains.

### `mutate()`: modify columns of the data

Use `mutate()` to add new columns to the data, or change/overwrite existing columns.

The general syntax inside `mutate()` is `new_col = function_of(old_col)`, or `old_col = function_of(old_col)` if you want to overwrite an existing column.

You can modify any kind of column: text data, numeric data, or otherwise, as long as you use the right functions that match that kind of data. For example, you can only `fct_recode()` on text columns, and you can only do math operations on numeric columns.

You can use many kinds of functions to modify these columns, as long as those functions work in a _one-to-one_ way: The function needs to be able to take a column as input, and return a column _of the same length_ as output.

```{r}
big5_modified <- big5_raw |> 
  mutate(gender_char = fct_recode(as.character(gender),
                                  "male" = "1",
                                  "female" = "2",
                                  "other" = "3",
                                  # I'm using this to set the missing values as NA
                                  NULL = "0"),
         hand_char = fct_recode(as.character(hand),
                                "right" = "1",
                                "left" = "2",
                                "both" = "3",
                                NULL = "0"),
         engnat_char = fct_recode(as.character(engnat),
                                  "yes" = "1",
                                  "no" = "2",
                                  NULL = "0"),
         across(.cols = c(E2, E4, E6, E8, E10, N2, N4, A1, A3, A5, A7, C2, C4, C6, C8, O2, O4, O6),
         .fns = \(x) 6-x)
         )
```

Using some special syntax, use the function `across()` inside of `mutate()` to apply the same function to multiple columns at once. By default, using `across()` will overwrite those columns in place, but there are more advanced argument settings you can specify to return new columns with different names.

### `filter()`: choose a subset of _rows_

Use `filter()` to return a shortened version of your dataframe, choosing rows based on relational statements with columns in the dataframe.

```{r}
big5_modified |> 
  filter(hand_char == "left")
```

Remember to spell column names and comparison terms correctly, otherwise you'll filter for the wrong thing!

Here are the relational operators you can use:

* `>` (greater than)
* `<` (less than)
* `>=` (greater than or equal to)
* `<=` (less than or equal to)
* `==` (is equal to--remember that it is _two_ equals signs)
* `!=` (is not equal to)
* `%in%` (is contained by; this is useful when you need to see whether the element on the left matches any of a vector of elements on the right)

To use `%in%`, you must specify the vector of potential values using `c()` to group them together:

```{r}
big5_modified |> 
  filter(hand_char %in% c("left", "both"))
```

You can modify/combine relational statements using these logical operators:

* `!` (NOT operator; this returns the opposite of whatever follows it)
* `&` (AND operator; this returns `TRUE` if both statements on either side are both true
* `|` (Pipe, or shift-backslash--OR operator; this returns `TRUE` if _at least one_ of the statements on either side is true)

If you use multiple relational statements one after the other separated by a comma, `filter()` combines them with **AND** by default.

```{r}
big5_modified |> 
  filter(hand_char %in% c("left", "both"),
         E1 > 3)
```

If you want to use logical **OR,** you must use the bar: `|`.

### `select()`: choose a subset of _columns_

Use `select()` to return a narrow version of your dataframe, choosing a subset of columns.

Generally, enter the column names you want to keep, _in the order you want them in the resulting dataframe._

```{r}
big5_modified |> 
  select(gender_char, hand_char, engnat_char, A1, A2, A3, A4, A5)
```

If you want to keep a range of adjacent columns, use the colon `:` to select "through". If you use the colon, you need to know what order the columns start off in so you know which columns land in the range!

```{r}
# Same result as the syntax above
# You can read that last bit as "A1 through A5"
big5_modified |> 
  select(gender_char, hand_char, engnat_char, A1:A5)
```

You can also use [helper functions](https://tidyselect.r-lib.org/reference/language.html) to select columns based on their names.

```{r}
big5_modified |> 
  select(ends_with("char"), A1:A5)
```

### `rename()`: rename columns

While you can use `select()` to rename columns, use `rename()` if the _only_ thing you want to do is rename columns (without dropping any columns or changing the column order).

They both use the syntax `new_col = old_col`. (When we `select()` without specifying a new column name, we are effectively just saying `old_col`.)

```{r}
big5_modified |> 
  # I looked, and I know E1 to O10 will capture all 50 personality columns
  select(ends_with("char"), E1:O10) |> 
  rename(gender = gender_char,
         hand = hand_char,
         engnat = engnat_char)
```

(Now that you've seen `rename()` in the lecture notes, I'll re-create the dataframe `big5_modified`, but renaming those columns and _saving them_ into the new object. Now, the new `big5_modified` will have the new column changes.)

```{r}
big5_modified <- big5_raw |> 
  mutate(gender_char = fct_recode(as.character(gender),
                                  "male" = "1",
                                  "female" = "2",
                                  "other" = "3",
                                  # I'm using this to set the missing values as NA
                                  NULL = "0"),
         hand_char = fct_recode(as.character(hand),
                                "right" = "1",
                                "left" = "2",
                                "both" = "3",
                                NULL = "0"),
         engnat_char = fct_recode(as.character(engnat),
                                  "yes" = "1",
                                  "no" = "2",
                                  NULL = "0"),
         across(.cols = c(E2, E4, E6, E8, E10, N2, N4, A1, A3, A5, A7, C2, C4, C6, C8, O2, O4, O6),
         .fns = \(x) 6-x)
         ) |> 
  select(ends_with("char"), E1:O10) |> 
  rename(gender = gender_char,
         hand = hand_char,
         engnat = engnat_char)
```


### `count()`: summarize by counting the number of rows

`count()` does one _specific_ type of summarizing: Counting the number of observations at each level of a categorical variable.

`count()` returns a _summarized_ dataframe. Instead of having one row per observation, you get one row per _level of the counted variable._

```{r}
big5_modified |> 
  count(engnat)
```

If you `count()` the levels of multiple variables at once, the `n` column gives you the observations at each possible combination of the counted variables:

```{r}
big5_modified |> 
  count(gender, engnat)
```

```{r, echo=FALSE}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "Food for thought: What happens when you try to count() the observations in a continuous variable? What does that tell you about how count() works? Try it for yourself and see!")
```

### `summarize()`: summarize using functions of your choice

`summarize()` is the general function for summarizing any columns in your data using any summary functions that satisfy the following rule:

A summarizing function must take _all_ the values in the column as inputs, and returns _one_ summary value as an output.

For example, we can use summary statistics functions like `mean()` and `sd()` to calculate the mean and standard deviation of some columns.

When we want to name our summarized output columns (which we almost always do), we use some familiar syntax: `summarized_col = function_of(input_col)`.

```{r}
big5_modified |> 
  summarize(mean_E1 = mean(E1),
            sd_E1 = sd(E1),
            mean_E2 = mean(E2),
            sd_E2 = sd(E2))
```

## Peek ahead

- **`select()` tricks:** How can we use `select()` to _drop_ columns (keep everything EXCEPT the columns specified)? Or _reorder_ columns without dropping any columns?
- **Summarizing for different groups:** How can you use the new function `group_by()` before `summarize()` so that instead of getting one summary value for the entire column, you get sub-summary values for _each level of a categorical variable?_ (For example, how can I get the mean value of E1 separately for men, women, and genderqueer participants, instead of across everyone?)
- **More knowledge through data manipulation:** How can you use `mutate()` to calculate each participant's _sum score_ for each of the Big 5 personality factors? For example, the total extraversion score is the sum of scores for E1 through E10. Once we have those sum scores, we can actually analyze patterns of personality scores across the dataset!
