---
title: "Overview"
output: blogdown::html_page
menu:
  docs:
    parent: "schedule"
weight: 200
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

We will _generally_ meet on **Mondays and Wednesdays 2:30-4:30 pm.** We may adjust meeting days some weeks depending on holidays, and to get some extra work time before final presentations.

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "The schedule is subject to change, but whatever you see here is current.")
```

## Week 1

### Monday

**Data science & scientific computing: what, why, and how?**

- What is data science?
- Why scientists program
- How to craft a data science research question

### Wednesday

**Sourcing useful data**

- Searching for datasets
- Evaluating the potential of a dataset
- The anatomy of code and code editors
- Reading in data

## Week 2

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "We will have class on Monday and Thursday this week (instead of Wednesday, for Juneteenth).")
```

**Data cleaning**

### Monday

- Tidy dataframe manipulation
- Data types
- Cleaning column formats

### _Thursday_

- Reshaping data
- Joining data
- Handling missing data

## Week 3

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "We will meet once on Thursday this week (Dr. Thieu is at a conference Mon-Wed).")
```

### _Thursday_

**Data cleaning (continued)**

- Additional data cleaning techniques

## Week 4

### Monday

**Exploratory data analysis**

- Exploratory graphs

### Wednesday

**Visual communication**

- Graphs as communication devices
- Visual communication best practices

## Week 5

### Monday

**Visual communication (continued)**

- Graph customization
- Critiquing graphs

### Wednesday

**Presentation design for clear communication**

- How to present data analysis projects
- Slide aesthetics
- Clarity and style in presentation design

## Week 6

**Completing projects**

### Monday

- Presentation work time

### Wednesday

- Presentation work time
- Publishing HTML presentations online

### _Thursday_

- Practice presentations

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "Presentations with all of WiSTEM on Friday!")
```
