---
title: "Overview"
output: blogdown::html_page
menu:
  docs:
    parent: "schedule"
weight: 200
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

We will _generally_ meet on **Mondays and Wednesdays 2:30-4:30 pm.** We may adjust meeting days some weeks depending on holidays, and to get some extra work time before final presentations.

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "The schedule is subject to change, but whatever you see here is current.")
```

## Week 1

### Monday June 10

**Data science & scientific computing: what, why, and how?**

- What is data science?
- Why scientists program
- How to craft a data science research question

### Wednesday June 13

**Sourcing useful data**

- Searching for datasets
- Evaluating the potential of a dataset
- The anatomy of code and code editors
- Reading in data

## Week 2

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "We will have class on Monday and Thursday this week (instead of Wednesday, for Juneteenth).")
```

**Data cleaning**

### Monday June 17

- Tidy dataframe manipulation
- Data types
- Cleaning column formats

### _Thursday June 20_

- Reshaping data
- Joining data
- Handling missing data

## Week 3

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "We will meet once on Thursday this week (Dr. Thieu is at a conference Mon-Wed).")
```

### _Thursday June 27_

**Data cleaning (continued)**

- Additional data cleaning techniques

## Week 4

### Monday July 1

**Exploratory data analysis**

- Exploratory graphs

### Wednesday July 3

**Visual communication**

- Graphs as communication devices
- Visual communication best practices

## Week 5

### Monday July 8

**Visual communication (continued)**

- Graph customization
- Critiquing graphs

### Wednesday July 10

**Presentation design for clear communication**

- How to present data analysis projects
- Slide aesthetics
- Clarity and style in presentation design

## Week 6

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "Different schedule this week! Dr. Thieu is traveling on Monday, and final presentations are Friday so we're meeting on Tues, Wed, and Thu.")
```

**Completing projects**

### _Tuesday July 16_

- Presentation work time

### Wednesday July 17

- Presentation work time
- Publishing HTML presentations online

### _Thursday July 18_

- Practice presentations

