---
title: "Overview"
output: blogdown::html_page
menu:
  docs:
    parent: "schedule"
weight: 200
toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

We will always meet at 2:30-4:30 pm. The weekly schedule will _generally_ be:

- Monday: Class
- Tuesday: _No meeting_
- Wednesday: Office hours and co-working
- Thursday: Class

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "The schedule is subject to change, but whatever you see here is current.")
```

## Week 1

### Monday

**Introduction to scientific computing**

- Why scientists program
- The anatomy of code and code editors
- How and why psychologists use computational tools

### Thursday

**Data wrangling: The core toolkit**

- Data types
- Logical and relational operators
- Tidy dataframe manipulation
- Troubleshooting as an experimentation skill

## Week 2

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "We will have a full class on _Tuesday,_ _Wednesday,_ and Thursday this week. We will not have class this Monday (Juneteenth).")
```

### _Tuesday_

**Data wrangling: The core toolkit (continued)**

- Tidy dataframe manipulation

### Wednesday

- Exploring data visually
- Dealing with irregular data
- Developing strategies for cleaning data

### Thursday

**Code style and hygiene**

- R scripts and R Markdown
- Scripting hygiene

## Week 3

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "We will have a slightly altered schedule this week (Dr. Thieu is at a conference).")
```

### _Tuesday_

- One-on-one check-in meetings (Zoom)

### Thursday

**Working with scientists' data**

- How to search for literature and datasets
- Inspecting and evaluating datasets
- Basic descriptive statistics

## Week 4

**Data cleaning: the real stuff**

### Monday

- Reading in data
- Cleaning column formats
- Handling missing data

### Thursday

- Reshaping data
- Joining data

## Week 5

**Graphing and visual communication**

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "We will have some additional lecture time this week to cover graphing and presentation styling techniques.")
```

### Monday

- Graphs as data analysis tools
- Exploratory data analysis

### Wednesday

- Graphs as communication devices
- Graph customization
- Visual communication best practices

### Thursday

- Slide aesthetics
- Clarity and style in presentation design

## Week 6

**Presenting and sharing code and results**

### Monday

- Presentation work time
- Publishing HTML write-ups online

### Wednesday

- How to present data analysis projects

### Thursday

- Practice presentations

```{r}
blogdown::shortcode_html(.name = "callout",
                         "note",
                         .content = "Presentations with all of WiSTEM on Friday!")
```
