---
title: 'Class notes: June 21, 2023'
output: blogdown::html_page
description: "Lecture notes on basic graphs with ggplot2."
excerpt: "Lecture notes on basic graphs with ggplot2."
date: 2023-06-21
lastmod: "`r Sys.Date()`"
draft: false
images: []
categories: ["Class notes"]
tags: ["2023"]
contributors: ["Monica Thieu"]
pinned: false
homepage: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Lecture notes

You can read more about today's topics in [Chapter 2 of R for Data Science](https://r4ds.hadley.nz/data-visualize.html).

First, we need to read in the data used for the demonstrations below.

```{r}
big5_raw <- read_tsv(here::here("static", "datasets", "BIG5_20140518", "data.csv"))
```

Then, we'll do some baseline data cleaning. In this data cleaning, we are:

- recoding the categorical variables
- reverse-scoring some of the individual personality questions
- using "special" `mutate()` with `rowwise()` to calculate means across each of the personality question columns
- using `select()` to keep only the recoded categorical columns and the personality sum columns
- using `rename()` to drop the "_char" suffix from the categorical columns without dropping any more columns

Notice that when we recode using `fct_recode()`, we can set the values originally coded as "0" to `NULL` in order to replace all the 0s with `NA`. We do this because we know from the codebook that 0 is the value used to indicate that the participant did not make a response. We want to replace those values with `NA` so that R knows they're _missing_ and will not include them in any stats.

```{r}
big5_modified <- big5_raw |> 
  mutate(gender_char = fct_recode(as.character(gender),
                                  "male" = "1",
                                  "female" = "2",
                                  "other" = "3",
                                  # I'm using this to set the missing values as NA
                                  NULL = "0"),
         hand_char = fct_recode(as.character(hand),
                                "right" = "1",
                                "left" = "2",
                                "both" = "3",
                                NULL = "0"),
         engnat_char = fct_recode(as.character(engnat),
                                  "yes" = "1",
                                  "no" = "2",
                                  NULL = "0"),
         across(.cols = c(E2, E4, E6, E8, E10, N2, N4, A1, A3, A5, A7, C2, C4, C6, C8, O2, O4, O6),
                .fns = \(x) 6-x)
  ) |> 
  rowwise() |> 
  mutate(C_sum = sum(c_across(C1:C10)),
         A_sum = sum(c_across(A1:A10)),
         N_sum = sum(c_across(N1:N10)),
         O_sum = sum(c_across(O1:O10)),
         E_sum = sum(c_across(E1:E10))) |> 
  ungroup() |> 
  select(ends_with("char"), ends_with("sum")) |> 
  rename(gender = gender_char,
         hand = hand_char,
         engnat = engnat_char)
```

### Graphing basics

In general, every `ggplot` command should follow this structure:

```r
# Start with the data you want to plot
dataframe_to_plot |> 
  # Create the "placemat" that goes under the plot
  ggplot(aes(x = desired_x_variable, y = desired_y_variable)) +
  # You can have multiple geom plots on top of one another if you want!
  geom_whatever() +
  # Always label and title the graph!
  labs(x = "Label on x axis",
       y = "Label on y axis",
       title = "Graph title")
```

Remember that ggplot2 commands must be combined with the `+`, not with the `|>`. Hopefully the developers will update the package to use the pipe soon!

#### Two continuous variables: scatter plots

A scatter plot is a great way to visualize the relationship between two continuous variables.

This plot shows the techniques we reviewed for improving the readability of scatter plots:

- using `geom_jitter()` will plot the points with a little bit of random noise to reduce _overplotting,_ or when too many points have the same value and plot right on top of one another
- reducing `alpha` inside the function will make the points less opaque/more transparent

```{r}
big5_modified |> 
  ggplot(aes(x = E_sum, y = A_sum)) +
  geom_jitter(alpha = 0.05) +
  labs(title = "Big 5 extraversion and agreeableness are slightly positively correlated",
       x = "Extraversion score (higher = more extraverted)",
       y = "Agreeableness score (higher = more agreeable)")
```

You can also add a line of best fit on top of the points to visualize the relationship between the x and y variables. (Like with everything, you can customize the colors of basically any element of your graph!)

```{r}
big5_modified |> 
  ggplot(aes(x = E_sum, y = A_sum)) +
  geom_jitter(alpha = 0.05) +
  geom_smooth(color = "orchid") +
  labs(title = "Big 5 extraversion and agreeableness are slightly positively correlated",
       x = "Extraversion score (higher = more extraverted)",
       y = "Agreeableness score (higher = more agreeable)")
```

#### One continuous variable: histogram

If you want to visualize the distribution of a single continuous variable, a histogram is the best way to do it. By plotting the count/height of observations at different values of X, you can see how the values are distributed.

It's important to set either the `bins` argument (number of histogram bins) or the `binwidth` argument (width of each bin, in units on the x-axis) to avoid some bins being weirdly taller than others because the bin cutoffs fall at values that don't make sense for the data. For example, because these personality scores can only be integers, I set `binwidth = 1` so that each score value gets its own bin.

```{r}
big5_modified |> 
  ggplot(aes(x = C_sum)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Conscientiousness score (higher = more conscientious)",
       title = "Conscientiousness is bell-shaped",
       subtitle ="BUT the curve peak is higher than 30/50, aka the middle score")
```


#### One continuous and one categorical variable: double histograms, box plots

In addition to visualizing the overall distribution of a continuous variable, you might want to know whether the distribution of that variable differs between observations at different levels of a second categorical variable.

One way to do that is with histograms like above, but with multiple histograms on top of one another for each of the groups of the categorical variable of interest.

Inside `aes()` in your initial `ggplot()` command, set `fill` to the categorical variable so that you can plot different histograms with different fill colors for each level of the categorical variable.

```{r}
big5_modified |> 
  ggplot(aes(x = C_sum, fill = gender)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.3) +
  labs(x = "Conscientiousness score (higher = more conscientious)",
       title = "Conscientiousness seems similarly distributed across genders")
```

Notice that histograms give you overall counts, so if one group has many more observations than another group (like how "female" is the largest category among these participants), its histogram will be taller overall.

To get a quick view of the distributions belonging to different levels of a categorical variable, you can use boxplots. Boxplots will show a box for the 25th-50th-75th percentiles, so that the middle 50% of observations fall inside the box, with the midline showing the median.

If your data have a sensible "middle" value (or any comparison value you want to mark on the plot), use `geom_hline()` for horizontal lines or `geom_vline()` for vertical lines. Note that the `yintercept` or `xintercept` argument goes right inside of `geom_hline()`/`geom_vline()`, when you want to set the line at a constant value (that doesn't depend on any data observations).

You can also control whether the line appears above or below your other plots by changing the order of the `geom` layers in your plot command.

```{r}
big5_modified |> 
  # filter only for the rows of non-missing gender
  # filter(!is.na(gender)) |> 
  ggplot(aes(x = gender, y = E_sum)) +
  geom_hline(yintercept = 30, linetype = "dotted", color = "#09a140") +
  geom_boxplot() +
  labs(y = "Conscientiousness score (higher = more conscientious)",
       title = "Conscientiousness by gender")
```

Keep exploring! If you want to get really into plots, the [official ggplot2 cookbook](https://r-graphics.org) has many different examples of plots to try.
